Constructing Statistical Deformable Models with Shape Flow

### 2015-07-16 contents/implementation.tex
updated sections related to shape flow theory

We propose a novel framework for building deformable models that does not require any consistent set of annotated landmarks and is based on a dense shape representation. Our method only requires a set of point or curve line annotations that does not need to be consistent over different training samples. It combines the techniques of Non-rigid Iterative Closest Point (NICP) \cite{Amber2007}, multi-channel Support Vector Shape (SVS) \cite{Nguyen2013} representation and multi-image subspace flow \cite{Garg:2013hu} in an effective framework that has significant descriptive power.

@@

In order to robustly establish dense correspondences across all SVS images using optical flow, it is important to constrain how pixels are allowed to move from one SVS image to another. In this work, we constrained the range of allowed pixels movements by learning a correspondence subspace.

In order to learn such a subspace, we first transform the original annotation to point clouds. Then, we use NICP to align the previous point clouds with respect to a reference point cloud template. NICP iteratively deforms each point cloud until its points match the ones on the shape template and correspondences are established. However, because optical flow is a pixel-wise frame registration technique, we need to establish correspondences for all pixels on the reference frame rather than for sparse point clouds. To achieve this, we apply Thin Plate Spline (TPS) \cite{Bookstein1989} to find correspondences for all pixels in the reference frame given the point cloud correspondences provided by the previous NICP stage. Once dense correspondences have been established among all pixels, the so-called correspondence subspace is obtained by performing PCA the trajectory of all pixels. Note that incorporating the previous subspace as a low rank constrained in optical flow is consistent with its assumption of smooth motion.

% For all dense transformations $\bm{u_n}(\bm{x}), n \in \{1,...,F\}$, where $F$ is number of data and $\bm{x}$ is vector of pixels, Principle Component Analysis (PCA) is performed on trajectory to obtain low rank trajectory basis:
% \begin{equation}
%     \begin{bmatrix}
%         \bm{u_1}(\bm{x}) \\
%         \vdots \\
%         \bm{u_F}(\bm{x})
%     \end{bmatrix}
%     =
%     \begin{bmatrix}
%         \bm{q_1}(1) & \cdots & \bm{q_R}(1) \\
%         \vdots      & \ddots & \vdots  \\
%         \bm{q_1}(F) & \cdots & \bm{q_R}(F)
%     \end{bmatrix}
%     \times
%     \begin{bmatrix}
%         \bm{v_1}(x) \\
%         \vdots \\
%         \bm{v_R}(x)
%     \end{bmatrix}
% \end{equation}
% where $\bm{q_i}(n)$ are low rank components with $R \ll 2F$ and $\bm{v_i}(x)$ weighted each component with dependencies on $x$. Simpler expression shown below:
% \begin{equation}
%     \bm{u_n}(\bm{x})=\sum_{i=1}^R\bm{q_i}(n)\bm{v_i}(x)+\bm{\varepsilon_n}(\bm{x})
% \end{equation}

@@

Using the SVS functions built from training data, the next step of our pipeline is to establish dense correspondences between all these functions. We propose to do this procedure jointly, by using the correspondence basis. This procedure shares similarities with multiframe optical flow and we call it Shape flow. We build upon robust methods for  optical flow estimation. Optical flow estimation typically works based on the assumptions of brightness or colour constancy and motion smoothness. However, in terms of shapes, neither constant illumination nor continuous motion exists in sequence of shapes as all training data are independent. 
For this reason, we propose to modify the formulation of multiframe optical flow by using the correspondence basis that we introduced in conjunction with the SVS functions, rather than actual images.

\subsubsection{Non-rigid ICP and Correspondence Basis Creation} \label{sec:trabasis}
Since shapes from training data set having completely no motion smoothness, it is of importance to contain constraints in optimization on how pixels moves from one shape frame to another. We constrain the optimization on correspondence subspace.
We supervise trajectories from sparse landmarks. To begin with, Non-rigid Iterative Closest Point (NICP)\cite{Amber2007} applied on sparse annotated points. NICP deforms target points to fit the template iteratively until all points on target deformed to match points on template. Performs the algorithm for all training data will get an implicit point correlation between sparse shapes.

Since optical flow pixel-wise frame registration, the trajectory basis build should also match the dimension, which is dense on shapes with trajectory length depends on number of training data. With correspondences produced after applying NICP, Thin-Plate Spline (TPS)\cite{Bookstein1989} transformations are utilized to warp every shape planes to template plane thereby generates an implicit dense shape registering. For all dense transformations $\bm{u_n}(\bm{x}), n \in \{1,...,F\}$, where $F$ is number of data and $\bm{x}$ is vector of pixels, Principle Component Analysis (PCA) is performed on trajectory to obtain low rank trajectory basis:
\begin{equation}
    \begin{bmatrix}
        \bm{u_1}(\bm{x}) \\
        \vdots \\
        \bm{u_F}(\bm{x})
    \end{bmatrix}
    =
    \begin{bmatrix}
        \bm{q_1}(1) & \cdots & \bm{q_R}(1) \\
        \vdots      & \ddots & \vdots  \\
        \bm{q_1}(F) & \cdots & \bm{q_R}(F)
    \end{bmatrix}
    \times
    \begin{bmatrix}
        \bm{v_1}(x) \\
        \vdots \\
        \bm{v_R}(x)
    \end{bmatrix}
\end{equation}
where $\bm{q_i}(n)$ are low rank components with $R \ll 2F$ and $\bm{v_i}(x)$ weighted each component with dependencies on $x$. Simpler expression shown below:
\begin{equation}
    \bm{u_n}(\bm{x})=\sum_{i=1}^R\bm{q_i}(n)\bm{v_i}(x)+\bm{\varepsilon_n}(\bm{x})
\end{equation}
Although optical flow on shapes made under the assumption that objects motion are smooth, the low rank constrain on trajectory basis recovered the hypothesis.

\subsubsection{Multi-image Subspace Flow}
To register all decision functions to template shape, decision function $d_i(\bm{x}), i \in {1,...,F}$ are grouped into one sequence before applying flow algorithm. The objective cost function we would like to minimise is:
\begin{align}
    \operatorname*{arg\,min}_{\bm{u_n}(\bm{x}), \bm{v}}&=\alpha \int_{\Omega}\sum_{n=1}^F|\bm{d_n}(x+\bm{u_n}(x))-\bm{d_0}(x)\| dx  \label{eq:costfunc}\\
    &+ \beta \int_{\Omega}\sum_{n=1}^F\|\bm{u_n}(x)-\sum_{i=1}^R\bm{q_i}(n)\bm{v_i}(x)\|^2 dx \label{eq:lowrank}\\
    &+ \sum[\bm{TV}(Qv)] + v^T.L.V
\end{align}
where $d_n(x)$ is the decision function from~\eqref{eq:decisionfunc}, which returns possibilities of given coordinate classified as shape component where coordinates are from set $\Omega \in \Re^2$. $TV(Qv)$ is total variation as regularization on low rank subspace, $Q$ is trajectory basis and $v^T.L.v$ is low rank spacial constraints.

Term~\eqref{eq:costfunc} state the shape constancy where points having similar classification probability are from same object.
Part~\eqref{eq:lowrank} applies constrain on low rank trajectory basis states in section~\ref{sec:trabasis}.
The objective cost function has two free parameters $u$ and $v$, so we perform alternating minimisation. The equation can be solved using a thresholding scheme after linearisation of image functions. The minimisation can be speed up by paralleling the minimisation for every spatial-temporal point $(x;n), x \in \Omega, n \in \{1,...F\}$ independently.

After solving the equation, $\bm{u_n}(\bm{x}), x \in \Omega$ gives a group of deformation fields that registering every decision functions in the shape sequence to reference frame e.g. $\bm{u_1}(\bm{x})$ registers decision function $\bm{d_1}(\bm{x})$ to the reference frame. Figure~\ref{fig:deformationfield} demonstrates deformation fields that warps from reference frame where there is no deformation.

>>>>

We propose a novel framework for building deformable models that does not require any consistent set of annotated landmarks and is based on a dense shape representation. Our method only requires a set of point or curve line annotations that does not need to be consistent over different training samples. It combines the techniques of Non-rigid Iterative Closest Point (NICP) \cite{Amber2007}, multi-channel Support Vector Shape (SVS) \cite{Nguyen2013} representation and multi-image subspace flow \cite{} in an effective framework that has significant descriptive power.

@@

We consider the problem of dense correspondences across SVS images as a joint estimation of optical flow fields between a reference SVS image and every SVS image of the training dataset. The estimation of such a ``shape flow'' defines for every training SVS image a warping function that registers it with the reference SVS image. To establish the dense correspondences robustly, we are inspired by the idea of subspace constraints in the estimation of multiframe optical flow \cite{Garg:2013hu}.

Instead of the \emph{motion} basis used in multiframe optical flow formulation of \cite{Garg:2013hu}, we build a \emph{correspondence} basis that introduces constraints on how points of different shapes are matched to each other. Every pixel of the reference SVS image is matched to its corresponding position at every training SVS image and in this way defines a \emph{correspondence vector}. This vector consists of the 2D locations of the specific point in all SVS images. To form this vector, the training images are arranged in an arbitrary order. Similarly to the order of training samples when Principal Component Analysis (PCA) is applied, this order does not affect the result of our method and any re-ordering would produce exactly the same results.

Formally, let $F$ be the number of training SVS images and $n=1,\ldots,F$ be the training image index. Also, let $\boldsymbol{q}_1(n),\ldots,\boldsymbol{q}_R(n):
\{1,\ldots,F\} \rightarrow \R^2$ be the $R$ orthonormal elements of the correspondence basis, where $\boldsymbol{q}_i(n)$ is the displacement vector that matches the points of the reference SVS image with the points of the $n$-th training SVS image, according to the variation described from the $i$-th correspondence basis element. Note that the basis elements $\boldsymbol{q}_i(n)$ are independent from the point location. Note also that the number of basis elements is typically much smaller than the full dimensionality ($2 F$) of correspondence vectors, therefore this basis plays a role of dimensionality reduction.

In addition, let $\Omega \subset \R^2$ be the image domain of the SVS images and $\bx$ denote the point location. We denote the shape flow result as $\bm{u}_n(\bx):\Omega\times \{1,\ldots,F\}
\rightarrow \R^2$,  where $\bm{u}_n(\bx)$ is the displacement vector that matches the point $\bx$ of the reference SVS image with its corresponding location at the $n$-th training SVS image.

Using the constructed correspondence basis, the shape flow can be approximated as:
\begin{equation}\label{eq:LinTrajModel}
    \bm{u}_n(\bx) \approx
    \sum_{i=1}^R\bm{q_i}(n)\bm{v_i}(\bx) \,\, ,
\end{equation}
where $\bm{v_i}(\bx)$ is the weight that needs to be applied on the $i$-th correspondence basis element, in order to get the correspondence vector for the point location $\bm{x}$. In other words, the shape flow for every point $\bm{x}$ is described as a linear combination of basis elements that is controlled by the coefficients $\bm{v_i}(\bm{x})$.
The values of the $i$-th coefficient for all the points $\bm{v_i}(\bm{x})$ can be interpreted as an image defined on $\Omega$. Using the correspondence basis, the determination of the shape flow boils down to the determination of the set of coefficients $\bm{v_i}(\bm{x})$. The above representation of shape flow constrains the correspondence vectors to lie on a subspace and therefore acts as a low-rank prior that enforces coherency of the shape registration result over the whole training dataset of shapes.

The next section describes the methodology of constructing the correspondence basis and estimating the shape flow robustly.

% For all dense transformations $\bm{u}_n(\bm{x}), n \in \{1,...,F\}$, where $F$ is number of data and $\bm{x}$ is vector of pixels, Principle Component Analysis (PCA) is performed on trajectory to obtain low rank trajectory basis:
% \begin{equation}
%     \begin{bmatrix}
%         \bm{u_1}(\bm{x}) \\
%         \vdots \\
%         \bm{u_F}(\bm{x})
%     \end{bmatrix}
%     =
%     \begin{bmatrix}
%         \bm{q_1}(1) & \cdots & \bm{q_R}(1) \\
%         \vdots      & \ddots & \vdots  \\
%         \bm{q_1}(F) & \cdots & \bm{q_R}(F)
%     \end{bmatrix}
%     \times
%     \begin{bmatrix}
%         \bm{v_1}(x) \\
%         \vdots \\
%         \bm{v_R}(x)
%     \end{bmatrix}
% \end{equation}
% where $\bm{q_i}(n)$ are low rank components with $R \ll 2F$ and $\bm{v_i}(x)$ weighted each component with dependencies on $x$. Simpler expression shown below:
% \begin{equation}
%     \bm{u}_n(\bm{x})=\sum_{i=1}^R\bm{q_i}(n)\bm{v_i}(x)+\bm{\varepsilon_n}(\bm{x})
% \end{equation}

@@

Using the SVS functions built from training data, one of the main steps of our pipeline is to estimate dense correspondences between all these functions. We propose to perform this estimation simultaneously for all training data, by using the correspondence basis. This procedure, which we call \emph{shape flow}, builds upon robust methods for multiframe optical flow estimation \cite{Garg:2013hu}. However, optical flow estimation typically works based on the assumptions of brightness or colour constancy and motion smoothness, whereas in our setting the input training data correspond to shapes. For this reason, we propose to modify the formulation of \cite{Garg:2013hu} by using the correspondence basis that we introduced in conjunction with the SVS representation of shapes.

%
%In order to learn the such a subspace, we first transform the original annotation to point clouds. Then, we use NICP to align the previous point clouds with respect to a reference point cloud template. NICP iteratively deforms each point cloud until its points match the ones on the shape template and correspondences are established. However, because optical flow is a pixel-wise frame registration technique, we need to establish correspondences for all pixels on the reference frame rather than for sparse point clouds. To achieve this, we apply Thin Plate Spline (TPS) \cite{Bookstein1989} to find correspondences for all pixels in the reference frame given the point cloud correspondences provided by the previous NICP stage. Once dense correspondences have been established among all pixels, the so-called correspondence subspace is obtained by performing PCA the trajectory of all pixels. Note that incorporating the previous subspace as a low rank constrained in optical flow is consistent with its assumption of smooth motion.

\subsubsection{Non-rigid ICP and Correspondence Basis Creation} \label{sec:trabasis}

To effectively build the correspondence basis, we first transform the original annotations to sparse point clouds. Then, we apply Non-rigid Iterative Closest Point (NICP)\cite{Amber2007} between the point cloud of annotations in the reference shape and the one of every shape of the training set. NICP iteratively deforms the cloud of points of every shape to match the points of the reference shape. This yields an initial estimation of the correspondence vectors on the sparse locations of annotated landmarks on the reference shape. To make this correspondence vectors estimation dense (i.e.~for every pixel of the reference SVS image), we apply Thin-Plate Splines (TPS)\cite{Bookstein1989} interpolation. Finally, the correspondence basis is found by applying PCA on the result of TPS interpolation and keeping only the first $R$ principal components.

%
%
%Since optical flow pixel-wise frame registration, the trajectory basis build should also match the dimension, which is dense on shapes with trajectory length depends on number of training data. With correspondences produced after applying NICP, Thin-Plate Spline (TPS)\cite{Bookstein1989} transformations are utilized to warp every shape planes to template plane thereby generates an implicit dense shape registering. For all dense transformations $\bm{u}_n(\bm{x}), n \in \{1,...,F\}$, where $F$ is number of data and $\bm{x}$ is vector of pixels, Principle Component Analysis (PCA) is performed on trajectory to obtain low rank trajectory basis:
%\begin{equation}
%    \begin{bmatrix}
%        \bm{u_1}(\bm{x}) \\
%        \vdots \\
%        \bm{u_F}(\bm{x})
%    \end{bmatrix}
%    =
%    \begin{bmatrix}
%        \bm{q_1}(1) & \cdots & \bm{q_R}(1) \\
%        \vdots      & \ddots & \vdots  \\
%        \bm{q_1}(F) & \cdots & \bm{q_R}(F)
%    \end{bmatrix}
%    \times
%    \begin{bmatrix}
%        \bm{v_1}(x) \\
%        \vdots \\
%        \bm{v_R}(x)
%    \end{bmatrix}
%\end{equation}
%where $\bm{q_i}(n)$ are low rank components with $R \ll 2F$ and $\bm{v_i}(x)$ weighted each component with dependencies on $x$. Simpler expression shown below:
%\begin{equation}
%    \bm{u}_n(\bm{x})=\sum_{i=1}^R\bm{q_i}(n)\bm{v_i}(\bm{x})+\bm{\varepsilon_n}(\bm{x})
%\end{equation}
%Although optical flow on shapes made under the assumption that objects motion are smooth, the low rank constrain on trajectory basis recovered the hypothesis.

\subsubsection{Variational Shape Flow Estimation}%{Multi-image Subspace Flow}

Following \cite{Garg:2013hu}, we propose to estimate the shape flow over all training SVS images $\bm{d_n}(\bx)$ by minimizing the following energy:
\begin{align}
E_{sf} & =\alpha
\int_{\Omega}\sum_{n=1}^F \|\bm{d_n}(\bx+\bm{u}_n(x))-\bm{d_0}(x)\| \ud \bx \label{eq:costfunc}\\
    &+ \beta \int_{\Omega}\sum_{n=1}^F\|\bm{u}_n(\bx)-\sum_{i=1}^R\bm{q_i}(n)\bm{v_i}(\bx)\|^2 \ud \bx \label{eq:lowrank}\\
    &+
\int_\Omega  \sum_{i=1}^R \,\, \left \|    \nabla \bm{v_i}(\bx)    \right \|  \,\ud \bx \label{eq:TVterm}
\end{align}
This energy consists of two sets of unknown shape flows that are relatively close to each other: (i) $\bm{u}_n(x)$, which tries to explain the data from the input SVS images, and (ii) the shape flow determined by the correspondence basis coefficients $\bm{v_i}(\bx)$ that are spatially regularised and enforce a low-rank prior. We minimize this energy jointly with respect to $\bm{u}_n(\bx)$ and $\bm{v_i}(\bx)$. The positive parameters $\alpha$ and $\beta$ weigh the balance between the terms of the energy.

The \textbf{first term} of the above energy (term \eqref{eq:costfunc}) is a data attachment term
that uses the robust $\Lone$-norm.  It is based on the assumption that the values of the reference SVS image $\bm{d_0}(\bx)$ at every pixel $\bx$ are preserved at its corresponding locations on all training SVS images $\bm{d_n}(\bx)$. The use of an $\Lone$-norm improves the robustness of the method since it allows deviations from this assumption, which might occur in practice.
The \textbf{second term} of the energy (term \eqref{eq:lowrank}) penalizes the difference between the two sets  of shape flows and acts as a coupling term between them.
The \textbf{third term} of the energy (term \eqref{eq:TVterm}) corresponds to the spatial Total Variation regularization \cite{rudin92} of
the correspondence basis coefficients $\bm{v_i}(\bx)$.
This term penalizes spatial oscillations of each coefficient caused by distortions of the SVS images but not strong discontinuities that are desirable in the borders of different object regions. In addition, this term allows to fill in information into regions where the shape information in the SVS images is missing, due to e.g.~regions with no annotations.

We implement the minimization of the energy $E_{sf}$ by using the optimization algorithm described in \cite{Garg:2013hu}. In every coarse-to-fine and warping iteration, we use an initialization that comes from the previous iteration and for the very beginning, we initialize with the shape flow result of the TPS interpolation described in Sec.~\ref{sec:trabasis}. We approximate the data term \eqref{eq:costfunc} by linearizing the SVS images around the initialization. After that, the energy becomes convex and we optimize it using an alternating optimization w.r.t.~$\bm{v_i}(\bx)$ and $\bm{u}_n(\bx)$. The minimization w.r.t.~
$\bm{v_i}(\bx)$ is decoupled for every coefficient $i$ and corresponds to Rudin-Osher-Fatemi Total Variation denoising \cite{rudin92}, which we solve efficiently by applying the first order primal-dual algorithm of \cite{Chambolle:Pock:JMIV2011}. The minimization w.r.t.~
$\bm{u}_n(x)$ is decoupled for every pixel $\bx$ and every shape index $i$. This minimization is also implemented by applying the efficient primal-dual algorithm of \cite{Chambolle:Pock:JMIV2011}.

Figure~\ref{fig:deformationfield} demonstrates some examples of the deformation fields derived by the estimated shape flow using the aforementioned method. These results correspond to some of the training shapes, in the case of a face dataset.

%
%
%To register all decision functions to template shape, decision function $d_i(\bm{x}), i \in {1,...,F}$ are grouped into one sequence before applying flow algorithm. The objective cost function we would like to minimise is:
%\begin{align}
%    \operatorname*{arg\,min}_{\bm{u}_n(\bm{x}), \bm{v}}&=\alpha \int_{\Omega}\sum_{n=1}^F|\bm{d_n}(x+\bm{u}_n(x))-\bm{d_0}(x)\| dx  \label{eq:costfunc}\\
%    &+ \beta \int_{\Omega}\sum_{n=1}^F\|\bm{u}_n(x)-\sum_{i=1}^R\bm{q_i}(n)\bm{v_i}(x)\|^2 dx \label{eq:lowrank}\\
%    &+ \sum[\bm{TV}(Qv)]
%\end{align}
%where $d_n(x)$ is the decision function from~\eqref{eq:decisionfunc}, which returns possibilities of given coordinate classified as shape component where coordinates are from set $\Omega \in \Re^2$. $TV(Qv)$ is total variation as regularization on low rank subspace, $Q$ is trajectory basis and $v^T.L.v$ is low rank spacial constrains.
%
%Term~\eqref{eq:costfunc} state the shape constancy where points having similar classification probability are from same object.
%Part~\eqref{eq:lowrank} applies constrain on low rank trajectory basis states in section~\ref{sec:trabasis}.
%The objective cost function has two free parameters $u$ and $v$, so we perform alternating minimisation. The equation can be solved using a thresholding scheme after linearisation of image functions. The minimisation can be speed up by paralleling the minimisation for every spatial-temporal point $(x;n), x \in \Omega, n \in \{1,...F\}$ independently.
%
%
%
%
%
%
%After solving the equation, $\bm{u}_n(\bm{x}), x \in \Omega$ gives a group of deformation fields that registering every decision functions in the shape sequence to reference frame e.g. $\bm{u_1}(\bm{x})$ registers decision function $\bm{d_1}(\bm{x})$ to the reference frame. Figure~\ref{fig:deformationfield} demonstrates deformation fields that warps from reference frame where there is no deformation.
